---
title: "AMI Program Implementation Assessment"
subtitle: "Water-use analysis, anomaly detection, and conservation targeting"
author: "Prepared with R + Snowflake"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    df_print: paged
    fig_width: 9
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))  # knit from project root
```

```{r load, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(lubridate)
  library(ggplot2); library(scales)
})
source("R/02_snowflake_connect.R")
source("R/03_anomaly_detection.R")
source("R/04_customer_classification.R")

reads  <- load_interval_reads(source = "auto")
meters <- load_meters()
theme_set(theme_minimal(base_size = 12))
pal <- c(normal = "#2c7fb8", leak = "#d7301f", continuous = "#fc8d59",
         stuck = "#7a0177", meter_error = "#000000", spike = "#fdae61")
```

## Executive summary

This assessment demonstrates an end-to-end Advanced Metering Infrastructure
(AMI) analysis workflow built in **R** on data managed in **Snowflake**. It
processes hourly interval reads to (1) characterize water-use patterns,
(2) automatically flag anomalies such as leaks and meter errors, (3) segment
customers for targeted conservation, and (4) quantify non-revenue water (NRW).

```{r summary-stats}
n_meters <- n_distinct(reads$meter_id)
n_reads  <- nrow(reads)
date_rng <- range(as_date(reads$read_ts))
total_gal <- sum(reads$consumption_gallons[reads$consumption_gallons > 0], na.rm = TRUE)

knitr::kable(
  tibble(
    Metric = c("Meters analyzed", "Hourly interval reads",
               "Period", "Total metered volume (gal)"),
    Value  = c(comma(n_meters), comma(n_reads),
               paste(date_rng, collapse = " to "), comma(round(total_gal)))
  ),
  caption = "Dataset at a glance"
)
```

> **Note on data.** This report runs on a realistic *synthetic* AMI dataset so
> it is fully reproducible without exposing client data. The identical workflow
> reads from a live Snowflake warehouse when credentials are configured
> (`source = "snowflake"`). See `README.md` for the data dictionary and the
> Snowflake DDL in `sql/`.

## 1. Water-use patterns

### System-wide daily demand

```{r daily-demand}
reads %>%
  mutate(read_date = as_date(read_ts)) %>%
  group_by(read_date) %>%
  summarise(gal = sum(consumption_gallons, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(read_date, gal)) +
  geom_line(color = "#2c7fb8", linewidth = 0.7) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Gallons / day",
       title = "System-wide daily metered demand",
       subtitle = "Seasonal ramp into summer reflects outdoor/irrigation demand")
```

### Diurnal profile by customer class

Different customer classes have distinct daily signatures. Residential demand is
twin-peaked (morning and evening); commercial accounts plateau through business
hours; irrigation concentrates in pre-dawn and evening watering windows. These
shapes drive both demand forecasting and anomaly baselines.

```{r diurnal}
reads %>%
  left_join(meters %>% select(meter_id, customer_class), by = "meter_id") %>%
  mutate(hr = hour(read_ts)) %>%
  group_by(customer_class, hr) %>%
  summarise(gal = mean(consumption_gallons, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(hr, gal, color = customer_class)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Hour of day", y = "Mean gallons / hour", color = NULL,
       title = "Average hourly use profile by customer class")
```

## 2. Anomaly detection

We run four transparent, parameterized detectors over the interval reads. Each
is designed to be **replicable across clients** by adjusting thresholds:

| Detector | Signal | Typical cause |
|---|---|---|
| **Leak** | Sustained step-up in overnight minimum flow (net of seasonal growth) | Underground/service-line leak, stuck fixture |
| **Continuous use** | Meter essentially never reads zero | Running toilet, stuck valve, process load |
| **Stuck meter** | Long run of identical/zero reads | Register failure, lost AMI comms, vacancy |
| **Meter error** | Negative or physically impossible reads | Register rollover, AMI decode error |

```{r detect}
anomalies <- run_all_detectors(reads) %>%
  left_join(meters, by = "meter_id")

anomalies %>%
  filter(any_anomaly) %>%
  count(primary_anomaly, name = "meters") %>%
  ggplot(aes(reorder(primary_anomaly, meters), meters, fill = primary_anomaly)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = pal) +
  coord_flip() +
  labs(x = NULL, y = "Meters flagged", title = "Flagged meters by anomaly type")
```

### Detector validation

Because the synthetic data carries known *ground-truth* anomalies, we can score
the detectors directly — evidence the methods actually work rather than just
producing plausible-looking output.

```{r validate}
gt_path <- "data/ground_truth_anomalies.csv"
if (file.exists(gt_path)) {
  ground_truth <- read_csv(gt_path, show_col_types = FALSE)
  scores <- score_detections(anomalies, ground_truth)
  knitr::kable(scores, digits = 2,
               caption = "Detection performance vs. injected ground truth")
} else {
  cat("Ground truth not available (live Snowflake mode).")
}
```

### Spatial view (GIS)

Mapping flagged meters by location helps spot geographic clustering — for
example, several leaks on one route can point to aging infrastructure in a
specific district-metered area. Meter coordinates here are synthetic.

```{r map}
map_df <- anomalies %>%
  mutate(status = if_else(any_anomaly, primary_anomaly, "normal"),
         sz = if_else(primary_anomaly == "leak", pmax(est_wasted_gallons, 1), 1))
ggplot() +
  geom_point(data = filter(map_df, !any_anomaly),
             aes(longitude, latitude), color = "grey75", size = 1.6, alpha = 0.7) +
  geom_point(data = filter(map_df, any_anomaly),
             aes(longitude, latitude, color = status, size = sz)) +
  scale_color_manual(values = pal, name = "Anomaly") +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  coord_quickmap() +
  labs(x = "Longitude", y = "Latitude",
       title = "Service-area map of flagged meters",
       subtitle = "Grey = normal; colored = anomalies (leaks sized by est. waste)")
```

> *Tip:* with the `leaflet` package installed, swap this static map for an
> interactive, zoomable web map — the dashboard includes the code.

### Highest-priority leaks

Estimated wasted volume helps prioritize field crews and customer outreach.

```{r leak-table}
anomalies %>%
  filter(primary_anomaly == "leak") %>%
  transmute(meter_id, customer_class, route_id,
            median_overnight_flow_gph = round(median_night_flow, 2),
            est_wasted_gallons = round(est_wasted_gallons)) %>%
  arrange(desc(est_wasted_gallons)) %>%
  head(10) %>%
  knitr::kable(caption = "Top leak candidates by estimated wasted volume")
```

## 3. Customer segmentation

Behavioural features (peakiness, overnight share, weekday/weekend ratio,
seasonal sensitivity) classify customers independent of their billing class —
portable across utilities. We show a transparent rule-based tiering plus an
unsupervised k-means discovery view.

```{r segments}
features <- build_usage_features(reads)
seg <- classify_rule_based(features) %>%
  left_join(meters %>% select(meter_id, customer_class), by = "meter_id")

seg %>%
  count(usage_tier, usage_pattern) %>%
  ggplot(aes(usage_tier, n, fill = usage_pattern)) +
  geom_col() +
  labs(x = "Usage tier", y = "Meters", fill = "Behavioural pattern",
       title = "Customer segments by usage tier and behaviour")
```

```{r kmeans}
km <- classify_kmeans(features, k = 4)
knitr::kable(km$profile, caption = "K-means cluster profiles (mean feature values)")
```

## 4. Non-revenue water (NRW)

NRW is the gap between water supplied into the system and water metered to
customers — a key efficiency and loss-control metric.

```{r nrw}
supply_path <- "data/system_supply.csv"
if (file.exists(supply_path)) {
  supply <- read_csv(supply_path, show_col_types = FALSE)
  daily_metered <- reads %>%
    mutate(supply_date = as_date(read_ts)) %>%
    group_by(supply_date) %>%
    summarise(metered_gallons = sum(consumption_gallons, na.rm = TRUE),
              .groups = "drop")
  nrw <- supply %>%
    mutate(supply_date = as_date(supply_date)) %>%
    inner_join(daily_metered, by = "supply_date") %>%
    mutate(nrw_percent = 100 * (supplied_gallons - metered_gallons) / supplied_gallons)

  ggplot(nrw, aes(supply_date, nrw_percent)) +
    geom_line(color = "#d7301f", linewidth = 0.6) +
    geom_hline(yintercept = mean(nrw$nrw_percent), linetype = "dashed") +
    labs(x = NULL, y = "Non-revenue water (%)",
         title = "Daily non-revenue water",
         subtitle = sprintf("Average %.1f%% over the period", mean(nrw$nrw_percent)))
}
```

## 5. Recommendations

Typical roadmap actions an analyst would draw from these results:

1. **Dispatch leak investigations** to the top-ranked overnight-flow meters,
   starting with the highest estimated wasted volume.
2. **Customer outreach** to continuous-use and high-overnight accounts (often a
   running toilet — a low-cost, high-impact fix).
3. **Data-quality remediation** for meters reporting negative/stuck reads before
   they corrupt billing and downstream analytics.
4. **Targeted conservation** to "Very High" tier and irrigation-driven segments,
   where per-account savings potential is largest.
5. **Track NRW** against the dashed baseline; sustained increases warrant a
   district-metered-area (DMA) loss audit.

---

*Reproduce this report:* run `R/01_generate_synthetic_ami.R` then knit this file
(`rmarkdown::render("reports/ami_assessment.Rmd")`), or point it at Snowflake by
configuring `config.yml` and setting `SNOWFLAKE_PASSWORD`.
